References: A Brief Overview

### 2023-04-25 3b_HowTos_LinearMixedModels.qmd
yaml section-nesting updates for how-tos

ggplot(df_long, aes(x = event, y = cbcl_extern)) + 
  geom_abline(model_slope = slopes2, model_intercept = intercepts2, 
              linetype = "solid", color = "black", linewidth = 1) + 
  # geom_abline(mapping = aes(slopes2 = model_slope, intercepts2 = model_intercept, linetype = PID), 
  #             linetype = "dashed", color = "grey70", linewidth = .4) +
  #geom_point(aes(shape = ids), size = 3.25, color = "grey70") + 
  #scale_shape_manual() + 
  # geom_segment(aes(x = event, xend = event, 
  #                  y = cbcl1, yend = fitted(model_slope)), 
  #              color = "grey70") +
  scale_y_continuous(expand = c(0, 0), breaks = c(0, 100, 200, 300, 400, 500), 
                     limits = c(0, 500)) +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9), 
                     limits = c(-0.5, 10.5)) +
  theme(panel.background = element_blank(),         
          panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.border = element_rect(colour = "black", fill = NA),
          legend.position = "none", 
          axis.text = element_text(size = 14),
          axis.title = element_text(size = 14)) +
    labs (x = "event", y = "cbcl_extern")

#Save the figure
 ggsave("random_slopes.png", units = "in", width = 9, height = 6, dpi = 300)
```

@@

```{r diag-plots}
#| messages: FALSE
#| warning: FALSE
#| echo: FALSE
#| column: screen-inset-shaded
#| layout-nrow: 2

## xxxxx plot
random_intercepts.std <- resid(model_intercept)/sd(resid(model_intercept))
plot (random_intercepts.std, ylab="Standardized Residuals")

## Outlier Tests. 
#car::outlierTest(random_intercepts)
car::infIndexPlot(random_intercepts)

## qqplot
ggplot(as.data.frame(random_intercepts.std), aes(sample = random_intercepts.std)) +
  geom_qq() +
  geom_qq_line()

## Predicted vs Actual plot
predicted <- estimate_expectation(random_intercepts, data = "grid")
plot(predicted)

## Estimated means plot
means <- estimate_means(random_intercepts)
plot(means)

@@

# Basic ggplot2 barplot with mean
p <-df_long %>%
    filter(!is.na(DBD)) %>% # filter on non-missing values  
    ggplot(aes(x = MatureVideoGames, y = Externalizing, fill = Sex)) +
    geom_col(position = "dodge",
           stat = "summary",
           fun = "mean",
           #fill = "#AA4A44",
           #color = "#0099f9"
           ) +
    #scale_fill_brewer(palette = "Set1") +
    scale_fill_manual(values = c("#3db5ff", "#0099f9")) +
    #scale_fill_gradient(high = "#B00B69", low = "#0e0e63") +
    #geom_text(aes(label = cu_traits), position = position_dodge(0.9), 
              #vjust = 2, size = 4, color = "#ffffff") +
    labs(title = "Behavior Problems by Video Game Groups", subtitle = "Simple bar chart",
         caption = "Simple caption", x = "DBD Group", y = "CBCL Externalizing") +
              #coord_flip() +
    #geom_hline(yintercept = mean(df$cbcl_agg), linetype = "dashed", size = 1) +
    theme_minimal() +
    theme(
          plot.title = element_text(color = "#0099f9", size = 20, hjust = 0.5),
          plot.subtitle = element_text(face = "bold", hjust = 1),
          plot.caption = element_text(face = "italic", hjust = 0),
            axis.title.x = element_text(color = "#0099f9", size = 15, face = "bold"),
            axis.title.y = element_text(size = 15, face = "italic")
    )
   p 
    ```

>>>>

# # ! THIS PLOT RUNS & PRINTS TO SCREEN. I am temporarily commenting it out due to a side-effect of the 
# #"intercepts2 variable"
# ggplot(df_long, aes(x = event, y = cbcl_extern)) + 
#   geom_abline(model_slope = slopes2, model_intercept = intercepts2, 
#               linetype = "solid", color = "black", linewidth = 1) + 
#   # geom_abline(mapping = aes(slopes2 = model_slope, intercepts2 = model_intercept, linetype = PID), 
#   #             linetype = "dashed", color = "grey70", linewidth = .4) +
#   #geom_point(aes(shape = ids), size = 3.25, color = "grey70") + 
#   #scale_shape_manual() + 
#   # geom_segment(aes(x = event, xend = event, 
#   #                  y = cbcl1, yend = fitted(model_slope)), 
#   #              color = "grey70") +
#   scale_y_continuous(expand = c(0, 0), breaks = c(0, 100, 200, 300, 400, 500), 
#                      limits = c(0, 500)) +
#   scale_x_continuous(expand = c(0, 0), breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9), 
#                      limits = c(-0.5, 10.5)) +
#   theme(panel.background = element_blank(),         
#           panel.grid.major = element_blank(),
#             panel.grid.minor = element_blank(),
#             panel.border = element_rect(colour = "black", fill = NA),
#           legend.position = "none", 
#           axis.text = element_text(size = 14),
#           axis.title = element_text(size = 14)) +
#     labs (x = "event", y = "cbcl_extern") 
#   
# #Save the figure
#  ggsave("random_slopes.png", units = "in", width = 9, height = 6, dpi = 300)
```

@@

```{r diag-plots}
# #| messages: FALSE
# #| warning: FALSE
# #| echo: FALSE
# #| column: screen-inset-shaded
# #| layout-nrow: 2
#     
# ## xxxxx plot
# random_intercepts.std <- resid(model_intercept)/sd(resid(model_intercept))
# plot (random_intercepts.std, ylab="Standardized Residuals")
# 
# ## Outlier Tests. 
# #car::outlierTest(random_intercepts)
# car::infIndexPlot(random_intercepts)
# 
# ## qqplot
# ggplot(as.data.frame(random_intercepts.std), aes(sample = random_intercepts.std)) +
#   geom_qq() +
#   geom_qq_line()
# 
# ## Predicted vs Actual plot
# predicted <- estimate_expectation(random_intercepts, data = "grid")
# plot(predicted)
# 
# ## Estimated means plot
# means <- estimate_means(random_intercepts)
# plot(means)

@@

# Basic ggplot2 barplot with mean
p <-df_long %>%
    filter(!is.na(event)) %>% # filter on non-missing values  
    ggplot(aes(x = mature_vg, y = cbcl_extern, fill = sex)) +
    geom_col(position = "dodge",
           stat = "summary",
           fun = "mean",
           #fill = "#AA4A44",
           #color = "#0099f9"
           ) +
    #scale_fill_brewer(palette = "Set1") +
    scale_fill_manual(values = c("#3db5ff", "#0099f9")) +
    #scale_fill_gradient(high = "#B00B69", low = "#0e0e63") +
    #geom_text(aes(label = cu_traits), position = position_dodge(0.9), 
              #vjust = 2, size = 4, color = "#ffffff") +
    labs(title = "Behavior Problems by Video Game Groups", subtitle = "Simple bar chart",
         caption = "Simple caption", x = "DBD Group", y = "CBCL Externalizing") +
              #coord_flip() +
    #geom_hline(yintercept = mean(df$cbcl_agg), linetype = "dashed", size = 1) +
    theme_minimal() +
    theme(
          plot.title = element_text(color = "#0099f9", size = 20, hjust = 0.5),
          plot.subtitle = element_text(face = "bold", hjust = 1),
          plot.caption = element_text(face = "italic", hjust = 0),
            axis.title.x = element_text(color = "#0099f9", size = 15, face = "bold"),
            axis.title.y = element_text(size = 15, face = "italic")
    )
```

### 2023-04-25 explanations.qmd
yaml section-nesting updates for how-tos

---
title: "Explanations: A Brief Overview"
number-sections: true
format:
 html:
   css: styles.css
   theme: cosmo
   fontsize: 1.1em
   linestretch: 1.5
---

## What are Explanations?

>>>>

---
title: "Explanations: A Brief Overview"
number-sections: false
format:
 html:
   css: styles.css
   theme: cosmo
   fontsize: 1.1em
   linestretch: 1.5
---

## ## <i class="bi bi-book"></i> What are Explanations?

### 2023-04-25 howto.qmd
yaml section-nesting updates for how-tos

---
title: "How-To's: A Brief Overview"
number-sections: true
format:
 html:
   css: styles.css
   theme: cosmo
   fontsize: 1.1em
   linestretch: 1.5
---

## What are How-To's?
::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.o; font-family: Georgia, serif;"}
[**The primary goal**]{style="color: goldenrod; font-size: 1.2em; font-family: Verdana, sans-serif;"} 
of the "how-to" section is on providing practical solutions to specific methods or analytic problems users might encounter. Unlike tutorials, which aim to teach a concept, how-to guides concentrate on the efficient execution of specific analyses without going into in-depth background explanations. By following these principles, the how-to section aims to create practical, efficient, and user-friendly guides that help users quickly identify solutions and accomplish their goals.
:::

>>>>

---
title: "How-To's: A Brief Overview"
number-sections: false
format:
 html:
   css: styles.css
   theme: cosmo
   fontsize: 1.1em
   linestretch: 1.5
---

## ## <i class="bi bi-book"></i>  What are How-To's?

::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.o; font-family: Georgia, serif;"}
[**The primary goal**]{style="color: goldenrod; font-size: 1.2em; font-family: Verdana, sans-serif;"} 
of the "how-to" section is on providing practical solutions to specific methods or analytic problems users might encounter. Unlike tutorials, which aim to teach a concept, how-to guides concentrate on the efficient execution of specific analyses without going into in-depth background explanations. By following these principles, the how-to section aims to create practical, efficient, and user-friendly guides that help users quickly identify solutions and accomplish their goals.
:::

### 2023-04-25 manuscript.qmd
yaml section-nesting updates for how-tos

---

title: Manuscript Working Draft

format: 
  pdf: 
    geometry: 
      - top=20mm
      - left=20mm
  docx: default
prefer-html: true

---

@@

# Part II Longitudinal Data: Interpretation / Issues / Pitfalls & Assumption
\label{sec:headings}
**Defining Features of Longitudinal Data Analysis.**
The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed. That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the “right” variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .

## Modeling Data Across Two Time Points versus Three or More Time Points.
Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, data based on two assessments are inherently limited on multiple fronts. As [@rogosa1982] noted approximately forty years ago, “Two waves of data are better than one, but maybe not much better”. These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for model identification and accurate parameter estimation. It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (since linear, straight line change is the only possible form, given two assessments; see [@duncan2009]). Research designs that include three or more time points allow for increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of inferential research. To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield20210603. It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability [e.g., the Latent Curve Model with Structured Residuals; @curran2014a] require at least four assessments to parameterize fully and, more generally, increasingly accurate parameter estimates are obtained as more assessment occasions are used [@duncan2009].

@@

---

### References

::: {#refs}
:::

>>>>

---
title: "Longitudinal Analysis Manuscript: Working Draft"
subtitle: Longitudinal Analysis using data from the ABCD Study 
authors:
  - name: Samuel W. Hawes
    department: Center for Children and Families
    affiliation: Florida International University
    location: Miami, FL
    email: shawes@fiu.edu
  - name: Andrew K. Littlefield
    department: Psychological Sciences
    affiliation: Texas Tech University
    location: Lubbock, TX
    email: andrew.littlefield@ttu.edu
  - name: Daniel A. Lopez
    department: Department of Public Health
    affiliation: University of Rochester Medical Center
    location: Rochester, NY
    email: Daniel_Lopez2@URMC.Rochester.edu
  - name: Kenneth J. Sher
    department: Psychological Sciences
    affiliation: University of Missouri
    location: Columbia, MO
    email: SherK@missouri.edu
  - name: Additional Co-authors
    department: xxxxxx
    affiliation: University of xxxx
    location: xxxxxxx
    email: xxxxxx
  - name: Wesley K. Thompson
    department: Center for Population Neuroscience and Genetics
    affiliation: Laureate Institute for Brain Research
    location: Tulsa, OK
    email: wes.stat@gmail.com

abstract: |
 The Adolescent Brain Cognitive Development (ABCD) Study presents a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of children and adolescents. Given the complex nature of the longitudinal data collected in the ABCD Study, researcher are likely to encounter a myriad of methodological and analytic considerations and concerns. This review provides a comprehensive examination of key issues and techniques related to longitudinal data analysis, specifically focusing on the ABCD Study. The text discusses model assumptions, common violations (e.g., independent and identically distributed residuals, *heterogeneous* *variability*) and their implications for valid inference. The importance of appropriately modeling covariance structures, understandings trade-offs between model fit and parsimony, and challenges related to sample size, attrition, missing data are highlighted. Consideration is given to the importance of selecting appropriate statistical models to account for correlations in repeated measurements and the assumptions underlying these models. The review also differentiates between linear and non-linear models in the context of continuous and discrete data, emphasizing various distributional assumptions and the necessity of choosing appropriate models and statistical methods. By addressing these complexities, the review seeks to equip researchers with the necessary knowledge and tools to make informed decisions as they navigate effectively analyzing and interpreting data available in the ABCD Study.

author-summary: |
  The Adolescent Brain Cognitive Development (ABCD) Study offers valuable insights into adolescent brain development. Researchers working with this data face methodological and analytic concerns, including sample size, attrition, missing data, and measurement. This review addresses complexities in analyzing longitudinal data, with a focus on statistical models accounting for correlations in repeated measurements and their assumptions. We emphasize the importance of choosing appropriate models and analytic functions based on the data and the specific scientific question to be addressed. This concise resource aims  to help researchers navigate longitudinal data analysis, make informed methodological decisions, and enhance the accuracy and reliability of their findings when working with data from the ABCD Study.

    keywords:
    *ABCD Study*,
    *longitudinal analyses*,
    *development*

format: 
  html:
    grid:
      sidebar-width: 200px
      margin-width: 300px
  pdf: 
    geometry:
      - top=30mm
      - left=20mm
    #fontfamily: libertinus
    #fontsize: 20pt
format-links: [html, pdf]

  #format:
#  pdf:
#    geometry:
#      - top=30mm
#      - left=20mm
#   docx: default
# prefer-html: true

output: 
  rticles::arxiv_article

bibliography: "references.bib" #references.bib
biblio-style: unsrt
---

@@

# Part II: Longitudinal Data: Interpretation / Issues / Pitfalls & Assumption
\label{sec:headings}
**Defining Features of Longitudinal Data Analysis.**
The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed. That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the “right” variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .

## Modeling Data Across Two Time Points versus Three or More Time Points.
Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, data based on two assessments are inherently limited on multiple fronts. As [@rogosa1982] noted approximately forty years ago, “Two waves of data are better than one, but maybe not much better”. These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for model identification and accurate parameter estimation. It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (since linear, straight line change is the only possible form, given two assessments; see [@duncan2009]). Research designs that include three or more time points allow for increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of inferential research. To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield2021. It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability [e.g., the Latent Curve Model with Structured Residuals; @curran2014a] require at least four assessments to parameterize fully and, more generally, increasingly accurate parameter estimates are obtained as more assessment occasions are used [@duncan2009].

@@

# Part III: Additional Section(s) in Progress: TBD]
\label{sec:headings}
**Subsection TBD.**

# Discussion
\label{sec:headings}
**Subsection TBD.**

### 2023-04-25 reference.qmd
yaml section-nesting updates for how-tos

---
title: "References: A Brief Overview"
number-sections: true
format:
 html:
   css: styles.css
   theme: cosmo
   fontsize: 1.1em
   linestretch: 1.5
---

## What are References?

>>>>

---
title: "References: A Brief Overview"
number-sections: false
format:
 html:
   css: styles.css
   theme: cosmo
   fontsize: 1.1em
   linestretch: 1.5
---

## ## <i class="bi bi-book"></i> What are References?
